Modeling trend in temperature volatility using generalized LASSO

### 2018-05-10 nips-2018/paper.tex
Porting to NIPS style

\documentclass[review]{elsarticle}

\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{}%
 \let\@evenfoot\@oddfoot}
\makeatother

\usepackage{etoolbox}
\patchcmd{\MaketitleBox}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{\today}\hfill\null\vfill}}{}{}%
\patchcmd{\pprintMaketitle}{\footnotesize\itshape\elsaddress\par\vskip36pt}{\footnotesize\itshape\elsaddress\par\parbox[b][36pt]{\linewidth}{\vfill\hfill\textnormal{Spring, 2018}\hfill\null\vfill}}{}{}%

% use Times
\usepackage{times}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}

% For citations
%\usepackage{natbib}

@@

\newcommand{\theHalgorithm}{\arabic{algorithm}}

@@

\bibliographystyle{model5-names}

@@

\begin{frontmatter}

\author{Arash Khodadadi}
\address{Department of Statistics, Indiana University, Bloomington, IN, United States}
\emailauthor{arash.khodadadi@gmail.com}{Arash Khodadadi}

\date{\today}

\begin{keyword}
generalized Lasso, linearized ADMM, consensus optimization, trend, volatility, convex optimization.
\end{keyword}

\end{frontmatter}

@@

There is a considerable interest in determining if there is an increasing trend in the climate variability \cite{hansen_perception_2012,huntingford_no_2013}. An increase in the temperature variability will increase the probability of extreme hot outliers. It might be harder for the society to adapt to these extremes than to the gradual increase in the mean temperature \cite{huntingford_no_2013}.

In this project, we consider the problem of detecting the trend in the temperature volatility. All analyses are performed on a sub-set of the European Centre for Medium-Range Weather Forecasts (ECMWF) ERA-40 dataset \cite{uppala_era-40_2005}. This dataset include the temperature measurements over a grid over the earth from 1957 to 2002.

Research on analyzing the trend in the volatility of spatio-temporal data is scarce. \cite{hansen_perception_2012} studied the change in the standard deviation (SD) of the surface temperature in the NASA Goddard Institute for Space Studies gridded temperature data set. In their analysis, for each geographical position, the mean of the temperature computed for the period 1951-1980 (called the base-period) at that position, is subtracted from the corresponding time series. Each time series is then divided by the standard deviation computed at each position and during the same time period. The distribution of the resulting data is then plotted for different periods. These distributions represent the deviation of the temperature for a specific period, from the mean in the base period, in units of the standard deviation in that period. The results showed that these distributions are widen for the resent time periods compared to 1951-1980.  \cite{huntingford_no_2013} took a similar approach in analysing the ERA-40 data set. However, in addition to the aforementioned method, they computed the distribution of the SDs in an alternative way: for each position and each time period, the deviation of the time-series at that position from the mean in that time period at that position was computed, and then divided by the SD of that position in the period before 1981. The results showed that there still is an increase in the SDs from 1958-1970 to 1991-2001, but this is much less than what is obtained from the method used in \cite{hansen_perception_2012}. The authors also computed the time-evolving global SD from the de-trended time-series at each position. The resulting curve suggested that the global SD has been stable.

These previous work (and other related research, e.g., \cite{rhines_frequent_2013}) have several shortcomings. First, no statistical analysis has been performed to examine if the change in the SD is statistically significant. Second, the methodologies for computing the SDs are rather arbitrary. The deviation of each time-series in a given period, is computed from either the mean of a base-period (as in \cite{hansen_perception_2012}), or from the given period (as in \cite{rhines_frequent_2013,huntingford_no_2013}). These deviations are then normalized using the SD of the base-period or the given period. No justification is provided for these choices. Third, the correlation between the observations is ignored. The observations in subsequent days and close geographical positions could be highly correlated. Without considering these correlations, any conclusion based on the averaged data could be flawed.

The main contribution of this work is to develop a new methodology for detecting the trend in the volatility of spatio-temporal data. In this methodology, the variance at each position and time, is considered as a hidden (un-observed) variable. The value of these hidden variables are then estimated by maximizing the likelihood of the observed data. We show that this formulation per se, is not appropriate for detecting the trend. To overcome this issue, we penalize the differences between the estimated variances of the observations which are temporally and/or spatially close to each other. This will result in an optimization problem called the \textit{generalized LASSO problem} \cite{tibshirani_solution_2011}. As we will see, the dimension of this optimization problem is very high and so the standard methods for solving the generalized LASSO cannot be applied directly. We investigate two methods for solving this optimization problem. In the first method, we adopt an optimization technique called alternative direction method of multipliers (ADMM) \cite{boyd_distributed_2011}, to divide the total problem into several sub-problems of much lower dimension and show how the total problem can be solved by iteratively solving these sub-problems. The second method, called the \textit{linearized ADMM algorithm} \cite{parikh_proximal_2014} solves the main problem by iteratively solving a linearized version of it. We will compare the benefits of each method.

@@

Figure \ref{fig:cities_ts} shows the time-series of the temperature of three cities: Bloomington (USA), San Diego (USA) and Manaus (Brazil). The time-series of Bloomington and San Diego show clear cyclic behavior. However, while it seems (qualitatively) that these cycles can be modeled by a sinusoidal function for Bloomington, the same is not true for San Diego. Also, the amplitude of the cycles changes from some years to others. The time-series of Manaus does not show any regular cyclic behavior. This demonstrates the first difficulty in analyzing the variance of this data: to analyze the variance, we first need to remove the cyclic terms from all time-series. However, there is a lot of variations in the cyclic behavior of the time-series of different locations. In addition, some of these cycles cannot be easily modeled by a parametric function \footnote{One might try to model the cycles by the summation of sinusoidal terms with different frequencies. However, for some time-series this may need many terms to be included in the summation to achieve a reasonable level of accuracy. In addition, this model cannot capture the non-stationarity in the cycles.}. To overcome these issues, we will use a non-parametric approach to remove the cyclic terms from the time-series and de-trend them. This approach, called \textit{$\ell_1$-trend filtering} is explained in the next section.

@@

The right panel of Figure \ref{fig:bloom_detrended} shows the time-series of the temperature of Bloomington, after removing the cyclic terms and de-trending using the method explained in the next section. The goal is to investigate the trend in the variance of this signal. This figure, reveals another issue toward this goal: the variance of this signal, shows cyclic behavior. Also, the cycles are not regular and their amplitude and frequency change. Even if one can describe the behavior of the variance of all the time-series using a single parametric model (for example a variant of the GARCH models \cite{bollerslev_generalized_1986}), it is not clear how the trend in the variance should be investigated in this framework. These observations motivate the need to develop a non-parametric framework for the problem at hand.

@@

$\ell_1$-trend filtering was proposed by \cite{kim_$ell_1$_2009} as a solution to the following problem: a time-series $y_t$, $t=1,...,T$ is observed and we believe that it consists of a smooth trend and a stochastic component and the goal is to estimate the trend. The estimated trend is desired to be smooth, but at the same time, we seek small estimated residuals (stochastic component). These two objectives cannot be achieved simultaneously and so a trade-off between them should be made. In $\ell_1$-trend filtering, this trade-off is formulated as the following convex optimization problem:

@@

\noindent where $||u||_1$ is the $\ell_1$ norm of the vector $u$. Here, $\beta$ is the vector of the free parameters of the model and there is one parameter per observation. The \textit{penalty matrix} is:

@@

The first term in Equation \ref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$. Specifically, with the penalty matrix \ref{eq:d_matrix}, the estimated $\beta$ will be piecewise linear \cite{kim_$ell_1$_2009}. The trade-off between the two aforementioned objectives is determined by the constant $\lambda$. \cite{kim_$ell_1$_2009} proposed a specialized primal-dual interior point (PDIP) method for solving the optimization problem \ref{eq:l1tf} with the penalty matrix \ref{eq:d_matrix}.

The results of applying $\ell_1$-trend filtering on the Bloomington time-series are shown in Figure \ref{fig:bloom_detrended}. The left panel shows the original time-series (blue) together with the estimated $\beta$ (red) and the right panel shows the residuals $y_t-\beta_t$ (blue). As it can be seen, the estimated $\beta$s are piecewise linear. The number of linear segments is determined by the value of $\lambda$. Here, we chose $\lambda=500$. The choice of $\lambda$ is a model selection problem and we will return to it in the next section ???.

@@

\noindent where $D$ is the same as in \ref{eq:d_matrix}.

@@

We solve this optimization problem using the PDIP method \footnote{We use PDIP implemented in the cvxopt Python package \cite{andersen_cvxopt:_2013}.}. First, we note that this is a generalized LASSO problem \cite{tibshirani_solution_2011}. The dual of a generalized LASSO with the objective $f(x)+\lambda \lVert Dx \lVert_1$ is:

@@

\noindent where $D^t$ is the transpose of $D$, and $f^*(\cdot)$ is the \textit{conjugate} of $f$ defined by: $f^*(u)=\max_x u^tx-f(x)$. It is simple to show that the conjugate of $f(h)=-l(y|h)$ is:

@@

The red curve in the left panel of Figure \ref{fig:bloom_estimatedSD} shows the estimated SD (which is $\exp(h_t/2)$) of the residuals of the time-series of Bloomington. To reduce the number of time-steps in this figure and in the remainder of the paper we work on the weekly averaged of the data.

The curve of the estimated SD captures the periodic variations in the SD of the signal. Just by looking at this curve, it is hard to say if the SD is decreasing or increasing. Therefore, we compute the average of the estimated SD for each year. The estimated SD together with this annual average is shown in the middle panel of Figure \ref{fig:bloom_estimatedSD}. As it can be seen, the annual trend is not smooth. This is because in the optimization problem \ref{eq:l1tf_var}, the smoothness of the annual trend is not encouraged. To remedy this, we add the following long horizon penalty to \ref{eq:l1tf_var}:

@@

\noindent where $t_1=52(i-1)+t$, $t_2=52i+t$ and $t_3=52(i+1)+t$. Also, $N_{year}$ is the number of years over which we are performing our analysis (here $N_{year}=10$). Since we are working on the weekly averaged data, each year corresponds to 52 observations. In the matrix form, the penalty \ref{eq:lh_penalty} adds $N_{year}$ rows to the matrix $D$. The estimated SDs using this penalty matrix is shown in the right panel of Figure \ref{fig:bloom_estimatedSD}. The annual average of the estimated SDs shows a linear trend with a positive slope.

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}}
		\caption{Left: The residuals of the time-series of Bloomington (averaged weekly) and the estimated SD obtained from the method of Section \ref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and their annual average (black) without imposing the long horizon penalty. Right: the same as middle panel but here the long horizon penalty is imposed. See the text for more details.}
		\label{fig:bloom_estimatedSD}
	\end{center}
	\vskip -0.2in
\end{figure}

@@

The first term in the objective is the negative log-likelihood (minus a constant term). The second term is the temporal penalty for the time-series at each location $(i,j)$. The third and fourth terms, penalize the difference between the estimated variance of two vertically and horizontally adjacent points, respectively. This penalty is a special case of the penalty used in the \textit{trend filtering on graphs }\cite{wang_trend_2016} (where the difference between the estimated values of the signal at each two nodes connected with an edge is penalized.). The optimization problem \ref{eq:l1tf_var_st} can be written in the matrix form. Let $h$ be a vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next $T$ entries are $h_{21t}$ and so on. Then the optimization problem in the matrix form is as follows:

@@

\noindent where $D$ was defined in \ref{eq:d_matrix}. The number of the diagonal blocks is equal to the grid size $n_r \times n_c$. Each row of the matrix $D_s$ corresponds to one spatial constraint in \ref{eq:l1tf_var_st}. For example, the first $T$ rows correspond to $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to $|h_{11t}-h_{12t}|$, and so on.

@@

\noindent where $f^*$ is given in \ref{eq:conj}.

\section{Optimization}
For a spatial grid of size $n_r \times n_c$ and for $T$ time steps, we have $n_t=3n_rn_c-Tn_c-2n_rn_c$ and $n_s=n_rn_cT$. For a grid over the united states and for weekly averaged data of 10 years we have $n_r=32$, $n_c=68$, $T=521$ and so $n_t\approx 3.5\times10^6$ and $n_s \approx 1.0\times10^6$. Therefore, the size of the optimization problem \ref{eq:dual} is about $n_t\approx 4.5\times10^6$. In each step of the PDIP algorithm, we need to solve a linear system of equations in the form $z=Ax$ where $A$ is a square matrix of size $2(n_t+n_s)$ (see \cite{boyd_convex_2004} equation 11.54). Therefore, applying the PDIP directly for solving the optimization problem \ref{eq:dual} is infeasible.

In the next section, we develop an ADMM algorithm for solving this problem. The idea is to cast the problem \ref{eq:dual} as a so-called \textit{consensus optimization problem} \cite{boyd_distributed_2011} and solve it by breaking the problem into smaller sub-problems using an ADMM-based algorithm. Next, we first give a brief overview of the consensus optimization problem and then explain how the problem \ref{eq:l1tf_var_st_mat} can be solved in this framework.

@@

A general form consensus optimization problem is in the form $\min_z \sum_i f_i(z)$, $z\in\mathbb{R}^n$. We can define a set of \textit{local variables} $x_i\in \mathbb{R}^{n_i}$ such that $\sum_i f_i(z)=\sum_i f_i(x_i)$. We follow the notation of \cite{boyd_distributed_2011} closely. Let $k=\mathscr{G}(i,j)$ which means that the $j^\text{th}$ entry of $x_i$ is $z_k$ (or $(x_i)_j=z_k$) and define $\tilde{z}_i \in \mathbb{R}^{n_i}$ by $(\tilde{z}_i)_j=(x_i)_j$. Then the original unconstrained optimization problem is equivalent to the following constrained optimization problem:

@@

To solve the optimization problem \ref{eq:l1tf_var_st_mat} or \ref{eq:l1tf_var_st} using the method explained in the previous section we need to two address two questions: first, how to choose the local variables $x_i$, and second, how to solve the optimization problem for updating these variables (the first line of \ref{eq:ADMM_steps}).

Obviously, the global variable in the problem \ref{eq:l1tf_var_st} is $z=h$. The global variable can be index by the spatial coordinate and the temporal step. In Figure \ref{fig:data_cube}, $z$ is represented as a cube. We can decompose $z$ into sub-cubes. An example of this decomposition is shown in the figure by white lines. It is easy to see that with this definition of $x_i$, the objective \ref{eq:l1tf_var_st_mat} decomposes as $\sum_i f_i(x_i)$ where $f_i(x_i)=-l(y_i|x_i)+\lVert \Lambda_i^t D_ix_i \lVert_1$.

@@

\noindent where:

@@

In these equations, $\mathscr{W}$ is the \textit{Lambert function} \cite{corless_lambertw_1996}. To use the PDIP method, we also need to compute the gradient and Hessian of $f_i^*(-D_i^t\nu_i)$. This involves computing the derivatives of the Lambert functions. The primal solution $x_i$ can be obtained from the dual solution $\nu_i$ in \ref{eq:dual_x_update} by setting $x_i=w_i$ where $w_i$ is defined in the last equation of \ref{eq:conj_x_update}.

The complete ADMM algorithm for estimating the variances is represented in \ref{alg:ADMM}. All the computations in the three updating steps \ref{eq:ADMM_steps} can be performed in parallel. The number of rows and columns of the sub-cubes should be chosen so that the updating of $x_i$ could be performed in one processor. We choose $3\times3\times521$ sub-cubes.

\begin{algorithm}[tb]
	\caption{ADMM for sparse estimation of variance of spatio-temporal data}
	\label{alg:ADMM}
	\begin{algorithmic}
		\STATE {\bfseries Input:} data $y$, mapping $\mathscr{G}(i,j)$, $\rho,\lambda_t,\lambda_s$
		\STATE Initialization: $x_i^0=z^0=u_i^0=\textbf{0}$.
		\FOR{$m=1,2,...$ }
			\FOR{$i=1$ {\bfseries to} $N_{sub-cubes}$}
				\STATE compute $\nu_i$ from \ref{eq:dual_x_update}
				\STATE compute $w_i$ from \ref{eq:conj_x_update}
				\STATE set $x_i^m:=w_i$
			\ENDFOR
		\STATE Compute $z^m$ from  \ref{eq:ADMM_steps}
		\STATE Compute $u_i^m$ from  \ref{eq:ADMM_steps}
		\ENDFOR
	\end{algorithmic}
\end{algorithm}

@@

In this section, we describe \textit{Linearized ADMM algorithm} \cite{parikh_proximal_2014} which, as we will see, makes the computation on a single machine feasible.

@@

\noindent where $x\in \mathbb{R}^n$ and $D\in \mathbb{R}^{m\times n}$. Each iteration of the linearized ADMM algorithm for solving this problem has the following form:

\begin{eqnarray}
\begin{aligned}
x^{k+1} & := \textbf{prox}_{\mu f} \big(x^k - (\mu/\rho)D^T (D x^k − z^k + u^k )\big)\\
z^{k+1} & := \textbf{prox}_{\rho g} \big(D x^k + u^k\big)\\
u^{k+1} & := u^k + D x^{k+1} - z^{k+1}
\end{aligned}
\label{eq:linADMM_steps}
\end{eqnarray}

\noindent where $z,u\in \mathbb{R}^m$ and the \textit{proximal operator} is defined as follows:

@@

This algorithm belongs to the general class of \textit{proximal algorithms} for solving convex optimization problems. For more details about these algorithms see \cite{parikh_proximal_2014}.

The optimization problem \ref{eq:l1tf_var_st} can be put into the form \ref{eq:linADMM_opt} as follows:

>>>>

\documentclass{article}
\PassOptionsToPackage{sort,numbers,square}{natbib}

\usepackage{nips_2018}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

% For citations
\usepackage{natbib}

@@

%\newcommand{\theHalgorithm}{\arabic{algorithm}}

\usepackage{graphicx}
\usepackage{subfigure} 
\graphicspath{{../working-paper/}}

@@

\usepackage[showonlyrefs]{mathtools}

\renewcommand*{\figureautorefname}{Figure}%
\renewcommand*{\tableautorefname}{Table}%
\renewcommand*{\sectionautorefname}{Section}%
\renewcommand*{\subsectionautorefname}{Section}%
\renewcommand*{\subsubsectionautorefname}{Section}%

\author{Arash Khodadadi\\
 Department of Statistics\\
 Indiana University\\
 Bloomington, IN 47408 \\
 \texttt{arakhoda@indiana.edu}
\And
  Daniel J.\ McDonald \\
 Department of Statistics\\
 Indiana University\\
 Bloomington, IN 47408 \\
 \texttt{dajmcdon@indiana.cmu.edu}}

@@

\maketitle

\begin{abstract}
words, words,

\end{abstract}

@@

There is a considerable interest in determining if there is an increasing trend in the climate variability \citep{hansen_perception_2012,huntingford_no_2013}. An increase in the temperature variability will increase the probability of extreme hot outliers. It might be harder for the society to adapt to these extremes than to the gradual increase in the mean temperature \citep{huntingford_no_2013}.

In this project, we consider the problem of detecting the trend in the temperature volatility. All analyses are performed on a sub-set of the European Centre for Medium-Range Weather Forecasts (ECMWF) ERA-40 dataset \citep{uppala_era-40_2005}. This dataset include the temperature measurements over a grid over the earth from 1957 to 2002.

Research on analyzing the trend in the volatility of spatio-temporal data is scarce. \citep{hansen_perception_2012} studied the change in the standard deviation (SD) of the surface temperature in the NASA Goddard Institute for Space Studies gridded temperature data set. In their analysis, for each geographical position, the mean of the temperature computed for the period 1951-1980 (called the base-period) at that position, is subtracted from the corresponding time series. Each time series is then divided by the standard deviation computed at each position and during the same time period. The distribution of the resulting data is then plotted for different periods. These distributions represent the deviation of the temperature for a specific period, from the mean in the base period, in units of the standard deviation in that period. The results showed that these distributions are widen for the resent time periods compared to 1951-1980.  \citep{huntingford_no_2013} took a similar approach in analysing the ERA-40 data set. However, in addition to the aforementioned method, they computed the distribution of the SDs in an alternative way: for each position and each time period, the deviation of the time-series at that position from the mean in that time period at that position was computed, and then divided by the SD of that position in the period before 1981. The results showed that there still is an increase in the SDs from 1958-1970 to 1991-2001, but this is much less than what is obtained from the method used in \citep{hansen_perception_2012}. The authors also computed the time-evolving global SD from the de-trended time-series at each position. The resulting curve suggested that the global SD has been stable.

These previous work (and other related research, e.g., \citep{rhines_frequent_2013}) have several shortcomings. First, no statistical analysis has been performed to examine if the change in the SD is statistically significant. Second, the methodologies for computing the SDs are rather arbitrary. The deviation of each time-series in a given period, is computed from either the mean of a base-period (as in \citep{hansen_perception_2012}), or from the given period (as in \citep{rhines_frequent_2013,huntingford_no_2013}). These deviations are then normalized using the SD of the base-period or the given period. No justification is provided for these choices. Third, the correlation between the observations is ignored. The observations in subsequent days and close geographical positions could be highly correlated. Without considering these correlations, any conclusion based on the averaged data could be flawed.

The main contribution of this work is to develop a new methodology for detecting the trend in the volatility of spatio-temporal data. In this methodology, the variance at each position and time, is considered as a hidden (un-observed) variable. The value of these hidden variables are then estimated by maximizing the likelihood of the observed data. We show that this formulation per se, is not appropriate for detecting the trend. To overcome this issue, we penalize the differences between the estimated variances of the observations which are temporally and/or spatially close to each other. This will result in an optimization problem called the \textit{generalized LASSO problem} \citep{tibshirani_solution_2011}. As we will see, the dimension of this optimization problem is very high and so the standard methods for solving the generalized LASSO cannot be applied directly. We investigate two methods for solving this optimization problem. In the first method, we adopt an optimization technique called alternative direction method of multipliers (ADMM) \citep{boyd_distributed_2011}, to divide the total problem into several sub-problems of much lower dimension and show how the total problem can be solved by iteratively solving these sub-problems. The second method, called the \textit{linearized ADMM algorithm} \citep{parikh_proximal_2014} solves the main problem by iteratively solving a linearized version of it. We will compare the benefits of each method.

@@

\autoref{fig:cities_ts} shows the time-series of the temperature of three cities: Bloomington (USA), San Diego (USA) and Manaus (Brazil). The time-series of Bloomington and San Diego show clear cyclic behavior. However, while it seems (qualitatively) that these cycles can be modeled by a sinusoidal function for Bloomington, the same is not true for San Diego. Also, the amplitude of the cycles changes from some years to others. The time-series of Manaus does not show any regular cyclic behavior. This demonstrates the first difficulty in analyzing the variance of this data: to analyze the variance, we first need to remove the cyclic terms from all time-series. However, there is a lot of variations in the cyclic behavior of the time-series of different locations. In addition, some of these cycles cannot be easily modeled by a parametric function \footnote{One might try to model the cycles by the summation of sinusoidal terms with different frequencies. However, for some time-series this may need many terms to be included in the summation to achieve a reasonable level of accuracy. In addition, this model cannot capture the non-stationarity in the cycles.}. To overcome these issues, we will use a non-parametric approach to remove the cyclic terms from the time-series and de-trend them. This approach, called \textit{$\ell_1$-trend filtering} is explained in the next section.

@@

The right panel of \autoref{fig:bloom_detrended} shows the time-series of the temperature of Bloomington, after removing the cyclic terms and de-trending using the method explained in the next section. The goal is to investigate the trend in the variance of this signal. This figure, reveals another issue toward this goal: the variance of this signal, shows cyclic behavior. Also, the cycles are not regular and their amplitude and frequency change. Even if one can describe the behavior of the variance of all the time-series using a single parametric model (for example a variant of the GARCH models \citep{bollerslev_generalized_1986}), it is not clear how the trend in the variance should be investigated in this framework. These observations motivate the need to develop a non-parametric framework for the problem at hand.

@@

$\ell_1$-trend filtering was proposed by \citep{kim_$ell_1$_2009} as a solution to the following problem: a time-series $y_t$, $t=1,...,T$ is observed and we believe that it consists of a smooth trend and a stochastic component and the goal is to estimate the trend. The estimated trend is desired to be smooth, but at the same time, we seek small estimated residuals (stochastic component). These two objectives cannot be achieved simultaneously and so a trade-off between them should be made. In $\ell_1$-trend filtering, this trade-off is formulated as the following convex optimization problem:

@@

 where $||u||_1$ is the $\ell_1$ norm of the vector $u$. Here, $\beta$ is the vector of the free parameters of the model and there is one parameter per observation. The \textit{penalty matrix} is:

@@

The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$. Specifically, with the penalty matrix \eqref{eq:d_matrix}, the estimated $\beta$ will be piecewise linear \citep{kim_$ell_1$_2009}. The trade-off between the two aforementioned objectives is determined by the constant $\lambda$. \citep{kim_$ell_1$_2009} proposed a specialized primal-dual interior point (PDIP) method for solving the optimization problem \eqref{eq:l1tf} with the penalty matrix \eqref{eq:d_matrix}.

The results of applying $\ell_1$-trend filtering on the Bloomington time-series are shown in \autoref{fig:bloom_detrended}. The left panel shows the original time-series (blue) together with the estimated $\beta$ (red) and the right panel shows the residuals $y_t-\beta_t$ (blue). As it can be seen, the estimated $\beta$s are piecewise linear. The number of linear segments is determined by the value of $\lambda$. Here, we chose $\lambda=500$. The choice of $\lambda$ is a model selection problem and we will return to it in the next section ???.

@@

 where $D$ is the same as in \eqref{eq:d_matrix}.

@@

We solve this optimization problem using the PDIP method \footnote{We use PDIP implemented in the cvxopt Python package \citep{andersen_cvxopt:_2013}.}. First, we note that this is a generalized LASSO problem \citep{tibshirani_solution_2011}. The dual of a generalized LASSO with the objective $f(x)+\lambda \lVert Dx \lVert_1$ is:

@@

 where $D^t$ is the transpose of $D$, and $f^*(\cdot)$ is the \textit{conjugate} of $f$ defined by: $f^*(u)=\max_x u^tx-f(x)$. It is simple to show that the conjugate of $f(h)=-l(y|h)$ is:

@@

The red curve in the left panel of \autoref{fig:bloom_estimatedSD} shows the estimated SD (which is $\exp(h_t/2)$) of the residuals of the time-series of Bloomington. To reduce the number of time-steps in this figure and in the remainder of the paper we work on the weekly averaged of the data.

The curve of the estimated SD captures the periodic variations in the SD of the signal. Just by looking at this curve, it is hard to say if the SD is decreasing or increasing. Therefore, we compute the average of the estimated SD for each year. The estimated SD together with this annual average is shown in the middle panel of \autoref{fig:bloom_estimatedSD}. As it can be seen, the annual trend is not smooth. This is because in the optimization problem \eqref{eq:l1tf_var}, the smoothness of the annual trend is not encouraged. To remedy this, we add the following long horizon penalty to \eqref{eq:l1tf_var}:

@@

 where $t_1=52(i-1)+t$, $t_2=52i+t$ and $t_3=52(i+1)+t$. Also, $N_{year}$ is the number of years over which we are performing our analysis (here $N_{year}=10$). Since we are working on the weekly averaged data, each year corresponds to 52 observations. In the matrix form, the penalty \eqref{eq:lh_penalty} adds $N_{year}$ rows to the matrix $D$. The estimated SDs using this penalty matrix is shown in the right panel of \autoref{fig:bloom_estimatedSD}. The annual average of the estimated SDs shows a linear trend with a positive slope.

\begin{figure}[ht]
	\vskip 0.2in
	\begin{center}
		\centerline{\includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}}
		\caption{Left: The residuals of the time-series of Bloomington (averaged weekly) and the estimated SD obtained from the method of \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and their annual average (black) without imposing the long horizon penalty. Right: the same as middle panel but here the long horizon penalty is imposed. See the text for more details.}
		\label{fig:bloom_estimatedSD}
	\end{center}
	\vskip -0.2in
\end{figure}

@@

The first term in the objective is the negative log-likelihood (minus a constant term). The second term is the temporal penalty for the time-series at each location $(i,j)$. The third and fourth terms, penalize the difference between the estimated variance of two vertically and horizontally adjacent points, respectively. This penalty is a special case of the penalty used in the \textit{trend filtering on graphs }\citep{wang_trend_2016} (where the difference between the estimated values of the signal at each two nodes connected with an edge is penalized.). The optimization problem \eqref{eq:l1tf_var_st} can be written in the matrix form. Let $h$ be a vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next $T$ entries are $h_{21t}$ and so on. Then the optimization problem in the matrix form is as follows:

@@

 where $D$ was defined in \eqref{eq:d_matrix}. The number of the diagonal blocks is equal to the grid size $n_r \times n_c$. Each row of the matrix $D_s$ corresponds to one spatial constraint in \eqref{eq:l1tf_var_st}. For example, the first $T$ rows correspond to $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to $|h_{11t}-h_{12t}|$, and so on.

@@

 where $f^*$ is given in \eqref{eq:conj}.

\section{Optimization}
For a spatial grid of size $n_r \times n_c$ and for $T$ time steps, we have $n_t=3n_rn_c-Tn_c-2n_rn_c$ and $n_s=n_rn_cT$. For a grid over the united states and for weekly averaged data of 10 years we have $n_r=32$, $n_c=68$, $T=521$ and so $n_t\approx 3.5\times10^6$ and $n_s \approx 1.0\times10^6$. Therefore, the size of the optimization problem \eqref{eq:dual} is about $n_t\approx 4.5\times10^6$. In each step of the PDIP algorithm, we need to solve a linear system of equations in the form $z=Ax$ where $A$ is a square matrix of size $2(n_t+n_s)$ (see \citep{boyd_convex_2004} equation 11.54). Therefore, applying the PDIP directly for solving the optimization problem \eqref{eq:dual} is infeasible.

In the next section, we develop an ADMM algorithm for solving this problem. The idea is to cast the problem \eqref{eq:dual} as a so-called \textit{consensus optimization problem} \citep{boyd_distributed_2011} and solve it by breaking the problem into smaller sub-problems using an ADMM-based algorithm. Next, we first give a brief overview of the consensus optimization problem and then explain how the problem \eqref{eq:l1tf_var_st_mat} can be solved in this framework.

@@

A general form consensus optimization problem is in the form $\min_z \sum_i f_i(z)$, $z\in\mathbb{R}^n$. We can define a set of \textit{local variables} $x_i\in \mathbb{R}^{n_i}$ such that $\sum_i f_i(z)=\sum_i f_i(x_i)$. We follow the notation of \citep{boyd_distributed_2011} closely. Let $k=\mathscr{G}(i,j)$ which means that the $j^\text{th}$ entry of $x_i$ is $z_k$ (or $(x_i)_j=z_k$) and define $\tilde{z}_i \in \mathbb{R}^{n_i}$ by $(\tilde{z}_i)_j=(x_i)_j$. Then the original unconstrained optimization problem is equivalent to the following constrained optimization problem:

@@

To solve the optimization problem \eqref{eq:l1tf_var_st_mat} or \eqref{eq:l1tf_var_st} using the method explained in the previous section we need to two address two questions: first, how to choose the local variables $x_i$, and second, how to solve the optimization problem for updating these variables (the first line of \eqref{eq:ADMM_steps}).

Obviously, the global variable in the problem \eqref{eq:l1tf_var_st} is $z=h$. The global variable can be index by the spatial coordinate and the temporal step. In \autoref{fig:data_cube}, $z$ is represented as a cube. We can decompose $z$ into sub-cubes. An example of this decomposition is shown in the figure by white lines. It is easy to see that with this definition of $x_i$, the objective \eqref{eq:l1tf_var_st_mat} decomposes as $\sum_i f_i(x_i)$ where $f_i(x_i)=-l(y_i|x_i)+\lVert \Lambda_i^t D_ix_i \lVert_1$.

@@

 where:

@@

In these equations, $\mathscr{W}$ is the \textit{Lambert function} \citep{corless_lambertw_1996}. To use the PDIP method, we also need to compute the gradient and Hessian of $f_i^*(-D_i^t\nu_i)$. This involves computing the derivatives of the Lambert functions. The primal solution $x_i$ can be obtained from the dual solution $\nu_i$ in \eqref{eq:dual_x_update} by setting $x_i=w_i$ where $w_i$ is defined in the last equation of \eqref{eq:conj_x_update}.

The complete ADMM algorithm for estimating the variances is represented in \autoref{alg:ADMM}. All the computations in the three updating steps \eqref{eq:ADMM_steps} can be performed in parallel. The number of rows and columns of the sub-cubes should be chosen so that the updating of $x_i$ could be performed in one processor. We choose $3\times3\times521$ sub-cubes.

\begin{algorithm}[tb]
	\caption{ADMM for sparse estimation of variance of spatio-temporal data}
	\label{alg:ADMM}
	\begin{algorithmic}
		\STATE {\bfseries Input:} data $y$, mapping $\mathscr{G}(i,j)$, $\rho,\lambda_t,\lambda_s$
		\STATE Initialization: $x_i^0=z^0=u_i^0=\textbf{0}$.
		\FOR{$m=1,2,...$ }
			\FOR{$i=1$ {\bfseries to} $N_{sub-cubes}$}
				\STATE compute $\nu_i$ from \eqref{eq:dual_x_update}
				\STATE compute $w_i$ from \eqref{eq:conj_x_update}
				\STATE set $x_i^m:=w_i$
			\ENDFOR
		\STATE Compute $z^m$ from  \eqref{eq:ADMM_steps}
		\STATE Compute $u_i^m$ from  \eqref{eq:ADMM_steps}
		\ENDFOR
	\end{algorithmic}
\end{algorithm}

@@

In this section, we describe \textit{Linearized ADMM algorithm} \citep{parikh_proximal_2014} which, as we will see, makes the computation on a single machine feasible.

@@

 where $x\in \mathbb{R}^n$ and $D\in \mathbb{R}^{m\times n}$. Each iteration of the linearized ADMM algorithm for solving this problem has the following form:

\begin{eqnarray}
\begin{aligned}
x^{k+1} & := \textbf{prox}_{\mu f} \big(x^k - (\mu/\rho)D^T (D x^k - z^k + u^k )\big)\\
z^{k+1} & := \textbf{prox}_{\rho g} \big(D x^k + u^k\big)\\
u^{k+1} & := u^k + D x^{k+1} - z^{k+1}
\end{aligned}
\label{eq:linADMM_steps}
\end{eqnarray}

 where $z,u\in \mathbb{R}^m$ and the \textit{proximal operator} is defined as follows:

@@

This algorithm belongs to the general class of \textit{proximal algorithms} for solving convex optimization problems. For more details about these algorithms see \citep{parikh_proximal_2014}.

The optimization problem \eqref{eq:l1tf_var_st} can be put into the form \eqref{eq:linADMM_opt} as follows:
